---
title: "Simulation"
author: "Hamed Nilforoshan"
date: "12/29/2021"
output:
  github_document:
    pandoc_args: --webtex
---


```{r setup, include=FALSE}
library(sigmoid)
library(tidyverse)
library(furrr)
library(BBmisc)
library(latex2exp)

# Set number of workers.
plan(multicore, workers = 4)

# Set the ggplot theme.
theme_set(theme_bw(base_size = 20))
```

## Overview

We construct a stylized example of college admissions to analyze the downstream consequences
of implementing many causal fairness definitions as decision-making constraints 

## Setup

A "book of life" is a draw of each potential outcome for a population of a given
size. We'll define a function that generates a book of life for a given set of
parameters. 

```{r book_of_life}

gen_book_of_life <- function(
  pop_size,
  frac_black,
  discrete,
  e_0,
  e_r, 
  e_noise,
  m_0,
  m_e,
  m_noise,
  t_0,
  t_m,
  t_e,
  t_e_m,
  t_noise,
  a_0,
  a_t,
  a_r,
  a_i,
  i_noise,
  b_p_reject_noise,
  b_p_reject_noise_mean,
  ...
) {
  
  # Define the structural equations.
  f_E <- function(r_i, noise){
    return(e_0 + e_r*r_i + noise)
  }
  
  f_M <- function(e_i, noise){
    return(m_0 + m_e*e_i + noise)
  }

  f_T <- function(m_i, e_i, noise){
  return(round(t_0 + t_m*m_i + t_e*e_i + t_e_m*e_i*m_i + noise ,digits=ifelse(discrete,0,3)))
  }
  
  f_A_raw <- function(t_i, r_i, i_i, a_i){
    return(a_t*t_i + a_r*r_i + a_0 + a_i*i_i)
  }
  
  f_A <- function(t_i, r_i, i_i, a_i){
    return(sigmoid(a_t*t_i + a_r*r_i + a_0 + a_i*i_i))
  }

  # Generate the book of life.
  book_of_life <- tibble(
      # Generate the exogenous variables.
      R = as.numeric(as.integer(rbernoulli(pop_size, frac_black))), # Race (Black=1, White=0)
      
      I_noise = 0,
      
      E_noise = rnorm(pop_size, mean = 0, sd = e_noise), #Educational opportunities
      E_black = f_E(1,noise = E_noise),
      E_white = f_E(0,noise = E_noise),
      
      M_noise = rnorm(pop_size, mean = 0, sd = m_noise), #Preparation
      M_black = f_M(E_black, noise=M_noise),
      M_white = f_M(E_white, noise=M_noise),

      # Generate the endogenous variables.
      T_noise = rnorm(pop_size, mean = 0, sd = t_noise), #Test score
      T_black = f_T(M_black, E_black, noise=T_noise),
      T_white = f_T(M_white, E_white, noise=T_noise),
      
      T_white_star = f_T(M_black, E_white, noise=T_noise),
      T_black_star = f_T(M_white, E_black, noise=T_noise),

      E = if_else(R==1,E_black,E_white),
      M = if_else(R==1,M_black,M_white),
      T = if_else(R==1,T_black,T_white),
      
      A_raw = f_A_raw(T, R,I_noise,a_i), # Admissions Committee Decision 
      A_raw_black = f_A_raw(T_black, 1,I_noise,a_i), # Admissions Committee Decision 
      A_raw_white = f_A_raw(T_white, 0,I_noise,a_i), # Admissions Committee Decision 

      A_prob = f_A(T,R,I_noise,a_i),
      A = as.integer(runif(pop_size,min=0,max=1) < A_prob),
      
      # Generate potential outcomes
      B_p_unif = runif(pop_size,min=0,max=1),
      B_p_reject_noise = rnorm(pop_size, mean = b_p_reject_noise_mean, sd = b_p_reject_noise),
      B_p = as.integer(sigmoid(M) >=B_p_unif) ,
      B_p_raw = sigmoid(M),
      B_p_reject = as.integer(sigmoid(M-0.5) >=B_p_unif),
      B_p_reject_raw = sigmoid(M-0.5),

      D_p = R , # Added diversity if admitted  
      
  )
}
```


## Initialize parameters and books of life
We initialize the training population size for the book of life, the fraction of the population that is a minority (black), and the number of counterfactual draws when calculating affirmative action decisions.

We then initial two books of life. One is for training our models, and the second is for testing what happens when the models are applied to a new, unseen dataset

```{r init}

set.seed(1)

POP_SIZE_TRAIN = 10000000
POP_SIZE_TEST = 100000
FRAC_BLACK = 0.33
DISCRETE = TRUE

E_0 = 1
E_R = -1
E_NOISE = 1
M_0 = -1.0
M_E = 1.0
M_NOISE = 1.0
T_0 = 50
T_M = 4
T_E = 4
T_E_M = 1.0
T_NOISE = 7
I_NOISE = 1.0
A_0 = -1.0

A_T = 0.02
A_R = -0.33
A_I = 0.33
B_P_REJECT_NOISE_SD = 0.3
B_P_REJECT_NOISE_MEAN = -0.5

df_train <-gen_book_of_life(pop_size=POP_SIZE_TRAIN,
                            frac_black=FRAC_BLACK,
                            discrete=DISCRETE,
                            e_0=E_0,
                            e_r=E_R,
                            e_noise=E_NOISE,
                            m_0=M_0,
                            m_e=M_E,
                            m_noise=M_NOISE,
                            t_0=T_0,
                            t_m=T_M,
                            t_e=T_E,
                            t_e_m=T_E_M,
                            t_noise=T_NOISE,
                            i_noise=I_NOISE,
                            a_0=A_0,
                            a_t=A_T,
                            a_r=A_R,
                            a_i=A_I,
                            b_p_reject_noise=B_P_REJECT_NOISE_SD,
                            b_p_reject_noise_mean=B_P_REJECT_NOISE_MEAN,)


df_test <-gen_book_of_life(pop_size=POP_SIZE_TEST,
                            frac_black=FRAC_BLACK,
                            discrete=DISCRETE,
                            e_0=E_0,
                            e_r=E_R,
                            e_noise=E_NOISE,
                            m_0=M_0,
                            m_e=M_E,
                            m_noise=M_NOISE,
                            t_0=T_0,
                            t_m=T_M,
                            t_e=T_E,
                            t_e_m=T_E_M,
                            t_noise=T_NOISE,
                            i_noise=I_NOISE,
                            a_0=A_0,
                            a_t=A_T,
                            a_r=A_R,
                            a_i=A_I,
                            b_p_reject_noise=B_P_REJECT_NOISE_SD,
                            b_p_reject_noise_mean=B_P_REJECT_NOISE_MEAN,)


DIVERSITY_UTILITY = 0.25
FRAC_ADMIT = 0.50 
POP_SIZE_TEST_ADMIT = FRAC_ADMIT * POP_SIZE_TEST #sum(df_test$A)

write.csv(df_train,'./df_train.csv')
write.csv(df_test,'./df_test.csv')
```



## Run linear programs in .ipynb and get output

```{r run_commands}

system(sprintf('export PATH="/Users/hamedn/opt/anaconda3/bin:$PATH"; 
        export DIVERSITY_UTILITY=%s;
        export FRAC_ADMIT=%s;
       jupyter nbconvert --execute --to notebook --inplace 00-Maximize-Utility.ipynb;
       jupyter nbconvert --execute --to notebook --inplace 01-Counterfactual-Fairness.ipynb;
       jupyter nbconvert --execute --to notebook --inplace 02-Path-Specific-Fairness.ipynb;
       jupyter nbconvert --execute --to notebook --inplace 03-Equalized_Odds.ipynb;
       jupyter nbconvert --execute --to notebook --inplace 04-Principal-Fairness.ipynb;
       jupyter nbconvert --execute --to notebook --inplace 05-Counterfactual-Predictive-Parity.ipynb;', DIVERSITY_UTILITY, FRAC_ADMIT))

```


```{r load_lp_results}
lp_results <- read.csv("lp_results.csv", sep = "\t")

PC_FAIR = c('Counterfactual Fairness\n& Path-Specific Fairness\n(Random)', 0, lp_results[lp_results$Policy.Name=='Counterfactual Fairness',]$Frac.Diverse.Candidates * POP_SIZE_TEST_ADMIT, lp_results[lp_results$Policy.Name=='Counterfactual Fairness',]$Total.Graduated)
EO_FAIR = c('Counterfactual Equalized Odds', 0, lp_results[lp_results$Policy.Name=='Counterfactual Equalized Odds',]$Frac.Diverse.Candidates *POP_SIZE_TEST_ADMIT , lp_results[lp_results$Policy.Name=='Counterfactual Equalized Odds',]$Total.Graduated)
PRINCIPAL_FAIR = c('Principal Fairness', 0, lp_results[lp_results$Policy.Name=='Principal Fairness',]$Frac.Diverse.Candidates *POP_SIZE_TEST_ADMIT , lp_results[lp_results$Policy.Name=='Principal Fairness',]$Total.Graduated)
ML_PREDICT_PAR = c('Counterfactual Pred. Parity', 0, lp_results[lp_results$Policy.Name=='Counterfactual Pred. Parity',]$Frac.Diverse.Candidates *POP_SIZE_TEST_ADMIT, lp_results[lp_results$Policy.Name=='Counterfactual Pred. Parity',]$Total.Graduated)

MAX_UTILITY = c(expression("Maximum utility (when " * lambda ~ " = 0.01)"), 0, lp_results[lp_results$Policy.Name=='Maximize Utility',]$Frac.Diverse.Candidates *POP_SIZE_TEST_ADMIT, lp_results[lp_results$Policy.Name=='Maximize Utility',]$Total.Graduated)

```




## Draw the pareto curve

```{r pareto_curve_02}
N=round(POP_SIZE_TEST_ADMIT)  * FRAC_BLACK * 2.0

MAX_BLACK_ADMIT = POP_SIZE_TEST_ADMIT * FRAC_BLACK* 2.0
pareto_ideal = NULL

for (num_black_admit in seq(from=1,to=FRAC_BLACK * POP_SIZE_TEST + 1000 ,by=as.integer(POP_SIZE_TEST_ADMIT/1000))){
  cat(num_black_admit)
  frac_black = num_black_admit

  df_test_black<- df_test[order(df_test$T,decreasing=TRUE),] %>% filter( R == 1 )
  df_test_white<- df_test[order(df_test$T,decreasing=TRUE),] %>% filter( R == 0 )
  lives_saved_outcomes = (sum(head(df_test_black,n=num_black_admit)$B_p) + sum(head(df_test_white,n=POP_SIZE_TEST_ADMIT-num_black_admit)$B_p))
  
  pareto_ideal = rbind(pareto_ideal, data.frame(frac_black ,lives_saved_outcomes))
}   

# ggplot colors
peach <- "#F8766D"
purple <- "#C77CFF"
blue <- "#00BFC4"
green <- "#00BE6C"
browngreen <- "#A3A500"
darkgreen <- "#DEAA88"
navy <- "#055C9D"
darkpurple <- "#431C53"

transparency <- 0.4
labels = c("1", "2", "3", "4", "5")

random_x <- 0.1
random_y <- 280000 

max_graduation <- pareto_ideal[which.max(pareto_ideal$lives_saved_outcomes),]
MAX_BOARDS_PASSAGE = c("Max Graduation", 0, max_graduation$frac_black, max_graduation$lives_saved_outcomes)

pareto_ideal$cutoff = pareto_ideal$frac_black >= as.double(MAX_BOARDS_PASSAGE[[3]])
lo = loess(lives_saved_outcomes ~ frac_black, pareto_ideal,span = 0.5 )
pareto_ideal$smooth_y = predict(lo, pareto_ideal$frac_black, se = FALSE)


p <- ggplot() +
  
  geom_line(data= filter(pareto_ideal, cutoff),aes(x=frac_black, y=smooth_y, color='ML Outcomes (Given T, R)'),size=1.5) + 
  geom_line(data= filter(pareto_ideal, !cutoff),aes(x=frac_black, y=smooth_y, color='ML Outcomes (Given T, R)'), linetype = "dashed",size=1.5) + 
  
  scale_x_continuous("Admitted Applicants from Target Group", labels = scales::comma) + 
  scale_y_continuous("College Degree Attainment", labels = scales::comma) +
  coord_cartesian(xlim = c(0.0, 30000),ylim = c(21800, 31500),expand = FALSE) + 
  geom_point(aes(x = as.double(MAX_BOARDS_PASSAGE[[3]]), y = as.double(MAX_BOARDS_PASSAGE[[4]])), color = peach, size = 4) +
  annotate("text",size=5, x = as.double(MAX_BOARDS_PASSAGE[[3]])-0.065-0.035-0.02+0.125+0.125, y = as.double(MAX_BOARDS_PASSAGE[[4]])+ (-200+250+250)/ 1.0, label = MAX_BOARDS_PASSAGE[[1]]) +

  geom_point(aes(x = as.double(MAX_UTILITY[[3]]), y = as.double(MAX_UTILITY[[4]])), color = blue, size = 4) +
  
  annotate("text",size=5, x = as.double(MAX_UTILITY[[3]])-0.06+0.075-0.01, y = as.double(MAX_UTILITY[[4]])-500/ 1.0, label = expression("Max Utility (for " * lambda ~  " = " * frac(1, 4) *")")) +

  geom_point(aes(x = as.double(PC_FAIR[[3]]), y = as.double(PC_FAIR[[4]])), color = purple, size = 4) +
  annotate("text",size=5,lineheight = 0.85,hjust = 1, x = as.double(PC_FAIR[[3]])-0.02, y = as.double(PC_FAIR[[4]]) +  500 + (-50+5000 + 100)/ POP_SIZE_TEST_ADMIT, label = PC_FAIR[[1]]) +
    
  geom_point(aes(x = as.double(PRINCIPAL_FAIR[[3]]), y = as.double(PRINCIPAL_FAIR[[4]])), color = darkgreen, size = 4) +
  annotate("text",size=5, x = as.double(PRINCIPAL_FAIR[[3]])-0.11-0.015+0.04, y = as.double(PRINCIPAL_FAIR[[4]]) +( - 50 + 100 - 500-100)/ POP_SIZE_TEST_ADMIT, label = PRINCIPAL_FAIR[[1]]) +

  geom_point(aes(x = as.double(ML_PREDICT_PAR[[3]]), y = as.double(ML_PREDICT_PAR[[4]])), color = green, size = 4) +
  annotate("text",size=5, x = as.double(ML_PREDICT_PAR[[3]])+0.13+0.02, y = as.double(ML_PREDICT_PAR[[4]]) + 50/ POP_SIZE_TEST_ADMIT, label = ML_PREDICT_PAR[[1]]) +
  
  geom_point(aes(x = as.double(EO_FAIR[[3]]), y = as.double(EO_FAIR[[4]])), color = browngreen, size = 4) +
  annotate("text",size=5,lineheight = 0.85,hjust = 1, x = as.double(EO_FAIR[[3]])-0.01, y = as.double(EO_FAIR[[4]]) +(- 750+200+80)/ POP_SIZE_TEST_ADMIT, label = EO_FAIR[[1]]) +

  scale_color_manual(
      values = c('ML Outcomes (Given T, R)'="violet" )) + 
  theme(legend.title = element_blank(),legend.position="none", axis.text=element_text(size=15), axis.title=element_text(size=18), panel.grid = element_line(color = rgb(235, 235, 235, 100, maxColorValue = 255)))
ggsave(plot=p, filename='./frontier.pdf', height=6, width=7)
p


```

