---
title: "Simulation"
author: "Hamed Nilforoshan"
date: "2/10/2021"
output:
  github_document:
    pandoc_args: --webtex
---


```{r setup, include=FALSE}
library(sigmoid)
library(tidymodels)
library(tidyverse)
library(furrr)

# Set number of workers.
plan(multicore, workers = 4)

# Set the ggplot theme.
theme_set(theme_bw(base_size = 20))
```

## Overview

We construct a stylized example illustrating the sub-optimal outcomes of Wang and Blei's fairness definitions (Affirmative Action and Equal Opportunity) on utility. 

## Setup

A "book of life" is a draw of each potential outcome for a population of a given
size. We'll define a function that generates a book of life for a given set of
parameters. T

```{r book_of_life}
gen_book_of_life <- function(
  pop_size,
  frac_black,
  lambda_1,
  lambda_2,
  lambda_3,
  beta_1,
  beta_2,
  beta_3,
  ...
) {
  
  # Define the structural equations.
  f_T <- function(r_i, m_i){
    return(round(pmax(0,pmin(lambda_1*r_i + lambda_2*m_i + rnorm(pop_size, mean = 0, sd = lambda_3) ,100)),digits=2))
  }
  
  f_A <- function(t_i, r_i){
    return(round(sigmoid(beta_1*t_i + beta_2*r_i + rnorm(pop_size, mean = 0, sd = beta_3) - 60.0), digits = 0))
  }
  

  # Generate the book of life.
  book_of_life <- tibble(
      # Generate the exogenous variables.
      R = as.integer(rbernoulli(pop_size, frac_black)), # Race (Black=1, White=0)
      M = runif(pop_size), # Medical Ability [0, 1]
      
      # Generate the endogenous variables.
      T_black = f_T(1,M), # Test Score black
      T_white = f_T(0,M), # Test Score black
      T = if_else(R==1,T_black,T_white),
      A = f_A(T, R), # Admissions Committee Decision 
      
      # Generate potential outcomes
      L_p = M * 100, # Lives saved if admitted
      D_p = R , # Added diversity if admitted  
      
      # Generate observed outcomes
      L_o = L_p * A, # Lives saved observed
      D_o = D_p * A  # Added diversity observed 
      
  )
}
```


## Initialize parameters and books of life
We initialize the training population size for the book of life, the fraction of the population that is a minority (black), and the number of counterfactual draws when calculating affirmative action decisions.

We then initial two books of life. One is for training our models, and the second is for testing what happens when the models are applied to a new, unseen dataset

```{r init}
POP_SIZE = 10000000
POP_SIZE_TEST = 500
POP_SIZE_TEST_ADMIT = 250
FRAC_BLACK = 0.2
COUNTER_FACTUAL_SAMPLE = 10000

df_train <-gen_book_of_life(POP_SIZE,FRAC_BLACK,-30,100,10,2,-25,5)
df_test <-gen_book_of_life(POP_SIZE_TEST,FRAC_BLACK,-30,100,10,2,-25,5)
```



## Use the biased admissions committee model as the baseline 

```{r ml_naive}

ml_naive <- glm(A ~ T + R, data = df_train, family = "binomial", control = list(maxit = 1000))
summary(ml_naive)

df_test$ml_naive = plogis(predict(ml_naive,df_test))
df_test<-df_test[order(df_test$ml_naive,decreasing=TRUE),]

sum(head(df_test,n=POP_SIZE_TEST_ADMIT)$L_p)
sum(head(df_test,n=POP_SIZE_TEST_ADMIT)$D_p)


summary(ml_naive)
```


## Use the outcomes model

```{r ml_outcomes}

df_train_ = filter(df_train, A == 1 )
ml_outcomes <- lm(L_p ~ T+R, data = df_train_)
summary(ml_outcomes)
df_test$ml_outcomes = predict(ml_outcomes,df_test)
df_test<-df_test[order(df_test$ml_outcomes,decreasing=TRUE),]
sum(head(df_test,n=POP_SIZE_TEST_ADMIT)$L_p)
sum(head(df_test,n=POP_SIZE_TEST_ADMIT)$D_p)

summary(ml_outcomes)
```


## Calculate the EO-Fair decision 

```{r ml_eo_fair}

df_test_black = tibble(df_test)
df_test_black$R=1
df_test_white = tibble(df_test)
df_test_white$R=0
df_test$ml_eo_fair = FRAC_BLACK*plogis(predict(ml_naive,df_test_black)) + (1-FRAC_BLACK)*plogis(predict(ml_naive,df_test_white))
df_test<-df_test[order(df_test$ml_eo_fair,decreasing=TRUE),]


sum(head(df_test,n=POP_SIZE_TEST_ADMIT)$L_p)
sum(head(df_test,n=POP_SIZE_TEST_ADMIT)$D_p)
```


## Calculate the AA-Fair decision

```{r aa_fair_model}
#Setup counterfactual probability distributions
aa_probs_black <- df_train %>%
  count(R, T, T_black) %>%
  group_by(R, T) %>%
  mutate(p = n / sum(n)) %>%
  ungroup()
  
aa_probs_white <- df_train %>%
  count(R, T, T_white) %>%
  group_by(R, T) %>%
  mutate(p = n / sum(n)) %>%
  ungroup()


AA_FAIR_MODEL <- function(applicant_race, applicant_test){
  
# Get sampling space for black and white counterfactuals
aa_probs_black_sample <- aa_probs_black %>% 
filter(R == applicant_race, T==applicant_test)
aa_probs_white_sample <- aa_probs_white %>% 
filter(R == applicant_race, T==applicant_test)

#sample test scores, first for black counterfactual, then for white
sampled_test_scores = aa_probs_black_sample[sample.int(nrow(aa_probs_black_sample), COUNTER_FACTUAL_SAMPLE*FRAC_BLACK, replace = TRUE, prob = aa_probs_black_sample$p),]$T_black
sampled_test_scores = c(sampled_test_scores,aa_probs_white_sample[sample.int(nrow(aa_probs_white_sample), COUNTER_FACTUAL_SAMPLE*(1-FRAC_BLACK), replace = TRUE, prob = aa_probs_white_sample$p),]$T_white)


aa_input_black <- tibble(R=1, T=sampled_test_scores)
aa_input_white <- tibble(R=0, T=sampled_test_scores)

AA_DECISION = mean(FRAC_BLACK*plogis(predict(ml_naive,aa_input_black)) + (1-FRAC_BLACK)*plogis(predict(ml_naive,aa_input_white)))

return(AA_DECISION)
  }
```



```{r ml_aaa_fair}
df_test <- df_test %>% 
  rowwise() %>% 
  mutate(ml_aa_fair = AA_FAIR_MODEL(R,T))


df_test<-df_test[order(df_test$ml_aa_fair,decreasing=TRUE),]
sum(head(df_test,n=POP_SIZE_TEST_ADMIT)$L_p)
sum(head(df_test,n=POP_SIZE_TEST_ADMIT)$D_p)
```

